Drought frequency predicts plant life history strategies

### 2018-11-28 manuscript/manuscript.tex
change csl

Understanding the causes and consequences of life history variation is a
longstanding goal of ecology and evolutionary biology (Cole, 1954). In
plants, life histories are especially diverse, with herbaceous species
that complete their life cycle in a number of weeks to trees that live
for thousands of years (Brown, 1996). Along this continuum an important
division exists, distinguishing annuals which complete their seed to
seed life cycle within a single calendar year from perennials which can
persist over multiple years. Annual plants flower once, set seed,
senesce, and then die, spending at least some portion of the year as a
seed, where they are relatively protected from environmental stress. In
contrast, perennial plants can continue vegetative growth after
reproduction and must survive conditions experienced during all seasons.
These represent fundamentally different life history strategies, but the
ecological factors that explain their evolution and distributions remain
empirically uresolved (Friedman \& Rubin, 2015).\\
Classical theory predict shorter life spans in environments where adult
mortality is high (Charnov \& Schaffer, 1973; Franco \& Silvertown,
1996; Stearns, 1992). In plants, this has been extended to the
hypothessis that annuality is adaptive when it allows plants to escape
drought (Schaffer \& Gadgil, 1975). Lack of water is perhaps the
greatest threat to survival during vegetatitive or reproductive growth
and annuals can remain dormant (and protected as a seed) during drought.
Thus, environments with greater seasonal drought frequency may select
for annual life histories that complete reproduction prior to drought
prone seasons. Conversely, environments with less frequent drought may
select for perennial species, which may benefit from multiple bouts of
reproduction and competitive advantage by preventing recruitment of
annual species (Corbin \& D'Antonio, 2004). These predictions have been
supported by the association of annualilty with arid environments in
\emph{Oryza perennis} (Morishima, Sano, \& Oka, 1984) and
\emph{Oenothera} (Evans, Hearn, Hahn, Spangle, \& Venable, 2005).
Additionally, annual and perennial species of \emph{Nemesia} were
qualitatively associated with winter rather and summer rainfall
environments respectively (Datson, Murray, \& Steiner, 2008) and annual
species of \emph{Scorzoneroides} were associated with environments
classified as unpredictable (Cruz-Mazo, Buide, Samuel, \& Narbona,
2009). However, whether the history frequency of drought events indeed
predicts the distributions annual or perennial life history strategies
has yet to be tested in a phylogenetic context with suitable sample
sizes of distrubution and climate data.\\
Here we combine a long-term global dataset of satellite detected drought
events with metadata from natural history collections to test these
classic hypotheses about the evolution of life history strategies within
the African endemic mustard genus, \emph{Heliophila} L. (Brassicaceae).
If annuallity is an adaptive strategy allowing plants to escape drought
prone seasons, then drought frequency should predict the distribution of
life history strategies across landscapes, and annual species should be
more commonly associated with drought prone regions than perennial
species. Furthermore, if annual species have adapted to escape drought
prone seasons, observations of growing annual species (i.e.~occurring in
forms other than seed) should be rare during drought prone seasons.
Phylogenetic relatedness can have significant non-random effects on
species distributions and life history traits (Barrett, Harder, \&
Worley, 1996), and therefore we assessed the relationship between life
history distribution and drought frequency in a phylogenetically
controlled background.

@@

Thus, VHI measurements are standardized according to conditions
historically observed at each locations. These measurements have been
validated and generally used for evaluating drought risk and predicting
crop yields in agriculture (e.g., Rojas, Vrieling, \& Rembold, 2011;
Kogan et al., 2016). But they also present a new tool to study seasonal
patterns in the frequency of drought across environments and to test
hypotheses about the effect of drought on ecological and evolutionary
processes (Kerr \& Ostrovsky, 2003). As such, the VHI has been applied
recently to study drought related ecology of natural species and proven
useful for predicting infraspecific variation in drought tolerance
traits and genes (Dittberner et al., 2018; Mojica et al., 2016; Monroe
et al., 2018). Here, we accessed VHI data at \(16km^2\) resolution from
1981 to 2015
(\url{https://www.star.nesdis.noaa.gov/smcd/emb/vci/VH/vh_ftp.php}) to
characterize the seasonal drought frequencies experienced by annual and
perennial \emph{Heliophila} species.

@@

\emph{Heliophila} is a genus of flowering plants endemic to the southern
portion of Africa including the Cape Floristic and Succulent Karoo
Regions. These are among the most botanically diverse environments on
Earth and the estimated \textasciitilde{}50 \emph{Heliophila} species
occurring there are considered to make up the most diverse genus of the
family Brassicaceae (Mandáková et al., 2012; Mummenhoff, Al-Shehbaz,
Bakker, Linder, \& Mühlhausen, 2005). This genus includes both perennial
and annual species and this change in life history strategy has likely
arisen multiple independent times (Appel \& Al-Shehbaz, 1997; Mummenhoff
et al., 2005). Furthermore, the fine scale climatic heterogeneity of
Southern Africa is ideal for studying the distribution of traits in
relation to environmental parameters (Sayre et al., 2013). We used life
histories reported by Mummenhoff et al. (2005), grouping species with
annual or perennial life histories. Perenniality was defined based any
form of perennial life history (e.g., herbs, shrubs, mixed, etc). We
also repeated these analyses without \emph{H. subulata} and \emph{H.
minima} which have mixed life histories and found similar results.
Because the original authors didn't report the nature of the mixed trait
(Mummenhoff et al., 2005), (ie. plasticity vs.~genetic variation), we
classified these species here as perennial since they have the capacity
to maintain vegetative growth after reproduction at least to some
capacity.

@@

Botanists have collected and maintained over 350 million botanical
specimens worldwide over the past 300 years. Herbarium specimens and
their associated metadata have been used since the 1960s to study
species' geographical distributions (reviewed by Willis et al. (2017)
and Lang, Willems, Scheepens, Burbano, and Bossdorf (2018)). And as they
become digitized (Soltis, 2017), these collections have been used to
study relationships between trait distributions, geography, and climate
(Davis, Willis, Connolly, Kelly, \& Ellison, 2015; Stropp et al., 2016;
Václavı'k, Beckmann, Cord, \& Bindewald, 2017; Wolf, Zimmerman,
Anderegg, Busby, \& Christensen, 2016). To characterize the diributions
of annual and perennial \emph{Heliophila} species, all records for the
genus \emph{Heliophila} were downloaded from the Global Biodiversity
Information Facility (gbif.org) on July 21, 2018 (GBIF, 2018).

@@

These findings support classical theoretical predictions about the
adaptive value of annual and perennial life history strategies. Taken
together, they suggest that in \emph{Heliophila}, annual species are
adapted to environments with increased summer droughts by avoiding these
seasons in a dormant seed phase of their life cycle. They also suggest
that perenniality is adaptive in environemnts where droughts are less
frequent. While most previous work has focused on describing the
evolutionary origins of annuality (Andreasen \& Baldwin, 2001; Barrett
et al., 1996; Conti, Soltis, Hardig, \& Schneider, 1999; Friedman \&
Rubin, 2015; Verboom, Linder, \& Stock, 2004) there are at least a few
other cases where perenniality appears to have arisen from an annual
acestor (Bena, Lejeune, Prosperi, \& Olivieri, 1998; Tank \& Olmstead,
2008). And while early theory predicted selection for annuality when
adult morality is high (Stearns, 1992), we also find evidence that the
transition to perenniality could be explained by historical drought
regimens. The phylogeny reveals several transitions from annual to
perennial life history (Figure \ref{fig:phylogeny}a) and that the
distributions of perennial \emph{Heliophila} extend into regions where
drought frequency is low (Figure \ref{fig:mapsboxplots}b,Figure
\ref{fig:speciesmaps}). Perennials may be able to out complete annual
relatives in environments where the infrequency of drought favors
strategies that allow plants to benefit from growth over many seasons.
It may also indicate that annuals rely on drought as a source of
disturbance for seedling recruitment when competing with perennials
(Corbin \& D'Antonio, 2004). Indeed, no annual species were observed in
the low drought regions of eastern South Africa
\ref{fig:mapsboxplots}a,Figure \ref{fig:speciesmaps}).

These findings suggest that species with locally adaptive life history
strategies could be threatened by rapidly changing drought regimens
(Dai, 2011). This could have impacts on ecosystem functioning and
processes such as carbon cycling if the composition of annual and
perennial species changes as a response {[}@garnier1997specific; Roumet,
Urcelay, and Dı'az (2006);monroe2018ecoevolutionary{]}. Furthermore, the
frequency of drought may be an important factor when considering the use
of perennial cropping sysyems (Lelièvre \& Volaire, 2009; Parry, Flexas,
\& Medrano, 2005).

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-aghakouchak2015remote}{}%
AghaKouchak, A., Farahmand, A., Melton, F., Teixeira, J., Anderson, M.,
Wardlow, B. D., \& Hain, C. (2015). Remote sensing of drought: Progress,
challenges and opportunities. \emph{Reviews of Geophysics},
\emph{53}(2), 452--480.

\leavevmode\hypertarget{ref-R-geiger_a}{}%
Alfaro, M., Santini, F., Brock, C., Alamillo, H., Dornburg, A., Rabosky,
D., \ldots{} Harmon, L. (2009). Nine exceptional radiations plus high
turnover explain species diversity in jawed vertebrates.
\emph{Proceedings of the National Academy of Sciences of the United
States of America}, \emph{106}, 13410--13414.

\leavevmode\hypertarget{ref-andreasen2001unequal}{}%
Andreasen, K., \& Baldwin, B. G. (2001). Unequal evolutionary rates
between annual and perennial lineages of checker mallows (sidalcea,
malvaceae): Evidence from 18S--26S rDNA internal and external
transcribed spacers. \emph{Molecular Biology and Evolution},
\emph{18}(6), 936--944.

\leavevmode\hypertarget{ref-appel1997generic}{}%
Appel, O., \& Al-Shehbaz, I. A. (1997). Generic limits and taxonomy of
hornungia, pritzelago, and hymenolobus (brassicaceae). \emph{Novon},
338--340.

\leavevmode\hypertarget{ref-R-papaja}{}%
Aust, F., \& Barth, M. (2018). \emph{papaja: Create APA manuscripts with
R Markdown}. Retrieved from \url{https://github.com/crsh/papaja}

\leavevmode\hypertarget{ref-barrett1996comparative}{}%
Barrett, S. C. H., Harder, L. D., \& Worley, A. C. (1996). The
comparative biology of pollination and mating in flowering plants.
\emph{Phil. Trans. R. Soc. Lond. B}, \emph{351}(1345), 1271--1280.

\leavevmode\hypertarget{ref-R-Matrix}{}%
Bates, D., \& Maechler, M. (2018). \emph{Matrix: Sparse and dense matrix
classes and methods}. Retrieved from
\url{https://CRAN.R-project.org/package=Matrix}

\leavevmode\hypertarget{ref-bena1998molecular}{}%
Bena, G., Lejeune, B., Prosperi, J.-M., \& Olivieri, I. (1998).
Molecular phylogenetic approach for studying life-history evolution: The
ambiguous example of the genus medicago l. \emph{Proceedings of the
Royal Society of London B: Biological Sciences}, \emph{265}(1401),
1141--1151.

\leavevmode\hypertarget{ref-brown1996oldlist}{}%
Brown, P. M. (1996). OLDLIST: A database of maximum tree ages.
\emph{Tree Rings, Environment, and Humanity. Radiocarbon}, \emph{1996},
727--731.

\leavevmode\hypertarget{ref-charnov1973life}{}%
Charnov, E. L., \& Schaffer, W. M. (1973). Life-history consequences of
natural selection: Cole's result revisited. \emph{The American
Naturalist}, \emph{107}(958), 791--793.

\leavevmode\hypertarget{ref-cole1954population}{}%
Cole, L. C. (1954). The population consequences of life history
phenomena. \emph{The Quarterly Review of Biology}, \emph{29}(2),
103--137.

\leavevmode\hypertarget{ref-conti1999phylogenetic}{}%
Conti, E., Soltis, D. E., Hardig, T. M., \& Schneider, J. (1999).
Phylogenetic relationships of the silver saxifrages (saxifraga, sect.
Ligulatae haworth): Implications for the evolution of substrate
specificity, life histories, and biogeography. \emph{Molecular
Phylogenetics and Evolution}, \emph{13}(3), 536--555.

\leavevmode\hypertarget{ref-corbin2004competition}{}%
Corbin, J. D., \& D'Antonio, C. M. (2004). Competition between native
perennial and exotic annual grasses: Implications for an historical
invasion. \emph{Ecology}, \emph{85}(5), 1273--1283.

\leavevmode\hypertarget{ref-cruz2009molecular}{}%
Cruz-Mazo, G., Buide, M., Samuel, R., \& Narbona, E. (2009). Molecular
phylogeny of scorzoneroides (asteraceae): Evolution of heterocarpy and
annual habit in unpredictable environments. \emph{Molecular
Phylogenetics and Evolution}, \emph{53}(3), 835--847.

\leavevmode\hypertarget{ref-dai2011drought}{}%
Dai, A. (2011). Drought under global warming: A review. \emph{Wiley
Interdisciplinary Reviews: Climate Change}, \emph{2}(1), 45--65.

\leavevmode\hypertarget{ref-daru2018widespread}{}%
Daru, B. H., Park, D. S., Primack, R. B., Willis, C. G., Barrington, D.
S., Whitfeld, T. J., \ldots{} others. (2018). Widespread sampling biases
in herbaria revealed from large-scale digitization. \emph{New
Phytologist}, \emph{217}(2), 939--955.

\leavevmode\hypertarget{ref-datson2008climate}{}%
Datson, P., Murray, B., \& Steiner, K. (2008). Climate and the evolution
of annual/perennial life-histories in nemesia (scrophulariaceae).
\emph{Plant Systematics and Evolution}, \emph{270}(1-2), 39--57.

\leavevmode\hypertarget{ref-davis2015herbarium}{}%
Davis, C. C., Willis, C. G., Connolly, B., Kelly, C., \& Ellison, A. M.
(2015). Herbarium records are reliable sources of phenological change
driven by climate and provide novel insights into species' phenological
cueing mechanisms. \emph{American Journal of Botany}, \emph{102}(10),
1599--1609.

\leavevmode\hypertarget{ref-dittberner2018natural}{}%
Dittberner, H., Korte, A., Mettler-Altmann, T., Weber, A., Monroe, G.,
\& Meaux, J. de. (2018). Natural variation in stomata size contributes
to the local adaptation of water-use efficiency in arabidopsis thaliana.
\emph{bioRxiv}, 253021.

\leavevmode\hypertarget{ref-R-geiger_b}{}%
Eastman, J., Alfaro, M., Joyce, P., Hipp, A., \& Harmon, L. (2011). A
novel comparative method for identifying shifts in the rate of character
evolution on trees. \emph{Evolution}, \emph{65}, 3578--3589.

\leavevmode\hypertarget{ref-evans2005climate}{}%
Evans, M. E., Hearn, D. J., Hahn, W. J., Spangle, J. M., \& Venable, D.
L. (2005). Climate and life-history evolution in evening primroses
(oenothera, onagraceae): A phylogenetic comparative analysis.
\emph{Evolution}, \emph{59}(9), 1914--1927.

\leavevmode\hypertarget{ref-franco1996life}{}%
Franco, M., \& Silvertown, J. (1996). Life history variation in plants:
An exploration of the fast-slow continuum hypothesis. \emph{Phil. Trans.
R. Soc. Lond. B}, \emph{351}(1345), 1341--1348.

\leavevmode\hypertarget{ref-friedman2015all}{}%
Friedman, J., \& Rubin, M. J. (2015). All in good time: Understanding
annual and perennial strategies in plants. \emph{American Journal of
Botany}, \emph{102}(4), 497--499.

\leavevmode\hypertarget{ref-gbifdownload}{}%
GBIF. (2018, July). GBIF occurrence download. Retrieved from
\url{https://doi.org/10.15468/dl.xmwqxm}

\leavevmode\hypertarget{ref-R-mvtnorm}{}%
Genz, A., \& Bretz, F. (2009). \emph{Computation of multivariate normal
and t probabilities}. Heidelberg: Springer-Verlag.

\leavevmode\hypertarget{ref-R-MCMCglmm}{}%
Hadfield, J. D. (2010). MCMC methods for multi-response generalized
linear mixed models: The MCMCglmm R package. \emph{Journal of
Statistical Software}, \emph{33}(2), 1--22. Retrieved from
\url{http://www.jstatsoft.org/v33/i02/}

\leavevmode\hypertarget{ref-R-geiger_d}{}%
Harmon, L., Weir, J., Brock, C., Glor, R., \& Challenger, W. (2008).
GEIGER: Investigating evolutionary radiations. \emph{Bioinformatics},
\emph{24}, 129--131.

\leavevmode\hypertarget{ref-R-logistf}{}%
Heinze, G., \& Ploner, M. (2018). \emph{Logistf: Firth's bias-reduced
logistic regression}. Retrieved from
\url{https://CRAN.R-project.org/package=logistf}

\leavevmode\hypertarget{ref-R-purrr}{}%
Henry, L., \& Wickham, H. (2018). \emph{Purrr: Functional programming
tools}. Retrieved from \url{https://CRAN.R-project.org/package=purrr}

\leavevmode\hypertarget{ref-R-raster}{}%
Hijmans, R. J. (2018). \emph{Raster: Geographic data analysis and
modeling}. Retrieved from
\url{https://CRAN.R-project.org/package=raster}

\leavevmode\hypertarget{ref-R-phylolm}{}%
Ho, L. S. T., \& Ane, C. (2014). A linear-time algorithm for gaussian
and non-gaussian trait evolution models. \emph{Systematic Biology},
\emph{63}, 397--408.

\leavevmode\hypertarget{ref-kerr2003space}{}%
Kerr, J. T., \& Ostrovsky, M. (2003). From space to species: Ecological
applications for remote sensing. \emph{Trends in Ecology \& Evolution},
\emph{18}(6), 299--305.

\leavevmode\hypertarget{ref-kogan2016modelling}{}%
Kogan, F., Guo, W., Strashnaia, A., Kleshenko, A., Chub, O., \&
Virchenko, O. (2016). Modelling and prediction of crop losses from noaa
polar-orbiting operational satellites. \emph{Geomatics, Natural Hazards
and Risk}, \emph{7}(3), 886--900.

\leavevmode\hypertarget{ref-kogan2001operational}{}%
Kogan, F. N. (2001). Operational space technology for global vegetation
assessment. \emph{Bulletin of the American Meteorological Society},
\emph{82}(9), 1949--1964.

>>>>

Understanding the causes and consequences of life history variation is a
longstanding goal of ecology and evolutionary biology (Cole, 1954). In
plants, life histories are especially diverse, with herbaceous species
that complete their life cycle in a number of weeks to trees that live
for thousands of years (Brown, 1996). Along this continuum an important
division exists, distinguishing annuals which complete their seed to
seed life cycle within a single calendar year from perennials which can
persist over multiple years. Annual plants flower once, set seed,
senesce, and then die, spending at least some portion of the year as a
seed, where they are relatively protected from environmental stress. In
contrast, perennial plants can continue vegetative growth after
reproduction and must survive conditions experienced during all seasons.
These represent fundamentally different life history strategies, but the
ecological factors that explain their evolution and distributions remain
empirically uresolved (Friedman and Rubin, 2015).\\
Classical theory predict shorter life spans in environments where adult
mortality is high (Charnov and Schaffer, 1973; Franco and Silvertown,
1996; Stearns, 1992). In plants, this has been extended to the
hypothessis that annuality is adaptive when it allows plants to escape
drought (Schaffer and Gadgil, 1975). Lack of water is perhaps the
greatest threat to survival during vegetatitive or reproductive growth
and annuals can remain dormant (and protected as a seed) during drought.
Thus, environments with greater seasonal drought frequency may select
for annual life histories that complete reproduction prior to drought
prone seasons. Conversely, environments with less frequent drought may
select for perennial species, which may benefit from multiple bouts of
reproduction and competitive advantage by preventing recruitment of
annual species (Corbin and D'Antonio, 2004). These predictions have been
supported by the association of annualilty with arid environments in
\emph{Oryza perennis} (Morishima et al., 1984) and \emph{Oenothera}
(Evans et al., 2005). Additionally, annual and perennial species of
\emph{Nemesia} were qualitatively associated with winter rather and
summer rainfall environments respectively (Datson et al., 2008) and
annual species of \emph{Scorzoneroides} were associated with
environments classified as unpredictable (Cruz-Mazo et al., 2009).
However, whether the history frequency of drought events indeed predicts
the distributions annual or perennial life history strategies has yet to
be tested in a phylogenetic context with suitable sample sizes of
distrubution and climate data.\\
Here we combine a long-term global dataset of satellite detected drought
events with metadata from natural history collections to test these
classic hypotheses about the evolution of life history strategies within
the African endemic mustard genus, \emph{Heliophila} L. (Brassicaceae).
If annuallity is an adaptive strategy allowing plants to escape drought
prone seasons, then drought frequency should predict the distribution of
life history strategies across landscapes, and annual species should be
more commonly associated with drought prone regions than perennial
species. Furthermore, if annual species have adapted to escape drought
prone seasons, observations of growing annual species (i.e.~occurring in
forms other than seed) should be rare during drought prone seasons.
Phylogenetic relatedness can have significant non-random effects on
species distributions and life history traits (Barrett et al., 1996),
and therefore we assessed the relationship between life history
distribution and drought frequency in a phylogenetically controlled
background.

@@

Thus, VHI measurements are standardized according to conditions
historically observed at each locations. These measurements have been
validated and generally used for evaluating drought risk and predicting
crop yields in agriculture (e.g., Rojas et al., 2011; Kogan et al.,
2016). But they also present a new tool to study seasonal patterns in
the frequency of drought across environments and to test hypotheses
about the effect of drought on ecological and evolutionary processes
(Kerr and Ostrovsky, 2003). As such, the VHI has been applied recently
to study drought related ecology of natural species and proven useful
for predicting infraspecific variation in drought tolerance traits and
genes (Dittberner et al., 2018; Mojica et al., 2016; Monroe et al.,
2018). Here, we accessed VHI data at \(16km^2\) resolution from 1981 to
2015 (\url{https://www.star.nesdis.noaa.gov/smcd/emb/vci/VH/vh_ftp.php})
to characterize the seasonal drought frequencies experienced by annual
and perennial \emph{Heliophila} species.

@@

\emph{Heliophila} is a genus of flowering plants endemic to the southern
portion of Africa including the Cape Floristic and Succulent Karoo
Regions. These are among the most botanically diverse environments on
Earth and the estimated \textasciitilde{}50 \emph{Heliophila} species
occurring there are considered to make up the most diverse genus of the
family Brassicaceae (Mandáková et al., 2012; Mummenhoff et al., 2005).
This genus includes both perennial and annual species and this change in
life history strategy has likely arisen multiple independent times
(Appel and Al-Shehbaz, 1997; Mummenhoff et al., 2005). Furthermore, the
fine scale climatic heterogeneity of Southern Africa is ideal for
studying the distribution of traits in relation to environmental
parameters (Sayre et al., 2013). We used life histories reported by
Mummenhoff et al. (2005), grouping species with annual or perennial life
histories. Perenniality was defined based any form of perennial life
history (e.g., herbs, shrubs, mixed, etc). We also repeated these
analyses without \emph{H. subulata} and \emph{H. minima} which have
mixed life histories and found similar results. Because the original
authors didn't report the nature of the mixed trait (Mummenhoff et al.,
2005), (ie. plasticity vs.~genetic variation), we classified these
species here as perennial since they have the capacity to maintain
vegetative growth after reproduction at least to some capacity.

@@

Botanists have collected and maintained over 350 million botanical
specimens worldwide over the past 300 years. Herbarium specimens and
their associated metadata have been used since the 1960s to study
species' geographical distributions (reviewed by Willis et al. (2017)
and Lang et al. (2018)). And as they become digitized (Soltis, 2017),
these collections have been used to study relationships between trait
distributions, geography, and climate (Davis et al., 2015; Stropp et
al., 2016; Václavı'k et al., 2017; Wolf et al., 2016). To characterize
the diributions of annual and perennial \emph{Heliophila} species, all
records for the genus \emph{Heliophila} were downloaded from the Global
Biodiversity Information Facility (gbif.org) on July 21, 2018 (GBIF,
2018).

@@

These findings support classical theoretical predictions about the
adaptive value of annual and perennial life history strategies. Taken
together, they suggest that in \emph{Heliophila}, annual species are
adapted to environments with increased summer droughts by avoiding these
seasons in a dormant seed phase of their life cycle. They also suggest
that perenniality is adaptive in environemnts where droughts are less
frequent. While most previous work has focused on describing the
evolutionary origins of annuality (Andreasen and Baldwin, 2001; Barrett
et al., 1996; Conti et al., 1999; Friedman and Rubin, 2015; Verboom et
al., 2004) there are at least a few other cases where perenniality
appears to have arisen from an annual acestor (Bena et al., 1998; Tank
and Olmstead, 2008). And while early theory predicted selection for
annuality when adult morality is high (Stearns, 1992), we also find
evidence that the transition to perenniality could be explained by
historical drought regimens. The phylogeny reveals several transitions
from annual to perennial life history (Figure \ref{fig:phylogeny}a) and
that the distributions of perennial \emph{Heliophila} extend into
regions where drought frequency is low (Figure
\ref{fig:mapsboxplots}b,Figure \ref{fig:speciesmaps}). Perennials may be
able to out complete annual relatives in environments where the
infrequency of drought favors strategies that allow plants to benefit
from growth over many seasons. It may also indicate that annuals rely on
drought as a source of disturbance for seedling recruitment when
competing with perennials (Corbin and D'Antonio, 2004). Indeed, no
annual species were observed in the low drought regions of eastern South
Africa \ref{fig:mapsboxplots}a,Figure \ref{fig:speciesmaps}).

These findings suggest that species with locally adaptive life history
strategies could be threatened by rapidly changing drought regimens
(Dai, 2011). This could have impacts on ecosystem functioning and
processes such as carbon cycling if the composition of annual and
perennial species changes as a response {[}@garnier1997specific; Roumet
et al. (2006);monroe2018ecoevolutionary{]}. Furthermore, the frequency
of drought may be an important factor when considering the use of
perennial cropping sysyems (Lelièvre and Volaire, 2009; Parry et al.,
2005).

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-aghakouchak2015remote}{}%
AghaKouchak A, Farahmand A, Melton F, Teixeira J, Anderson M, Wardlow
BD, Hain C. 2015. Remote sensing of drought: Progress, challenges and
opportunities. \emph{Reviews of Geophysics} \textbf{53}:452--480.

\leavevmode\hypertarget{ref-R-geiger_a}{}%
Alfaro M, Santini F, Brock C, Alamillo H, Dornburg A, Rabosky D,
Carnevale G, Harmon L. 2009. Nine exceptional radiations plus high
turnover explain species diversity in jawed vertebrates.
\emph{Proceedings of the National Academy of Sciences of the United
States of America} \textbf{106}:13410--13414.

\leavevmode\hypertarget{ref-andreasen2001unequal}{}%
Andreasen K, Baldwin BG. 2001. Unequal evolutionary rates between annual
and perennial lineages of checker mallows (sidalcea, malvaceae):
Evidence from 18S--26S rDNA internal and external transcribed spacers.
\emph{Molecular Biology and Evolution} \textbf{18}:936--944.

\leavevmode\hypertarget{ref-appel1997generic}{}%
Appel O, Al-Shehbaz IA. 1997. Generic limits and taxonomy of hornungia,
pritzelago, and hymenolobus (brassicaceae). \emph{Novon} 338--340.

\leavevmode\hypertarget{ref-R-papaja}{}%
Aust F, Barth M. 2018. papaja: Create APA manuscripts with R Markdown.

\leavevmode\hypertarget{ref-barrett1996comparative}{}%
Barrett SCH, Harder LD, Worley AC. 1996. The comparative biology of
pollination and mating in flowering plants. \emph{Phil Trans R Soc Lond
B} \textbf{351}:1271--1280.

\leavevmode\hypertarget{ref-R-Matrix}{}%
Bates D, Maechler M. 2018. Matrix: Sparse and dense matrix classes and
methods.

\leavevmode\hypertarget{ref-bena1998molecular}{}%
Bena G, Lejeune B, Prosperi J-M, Olivieri I. 1998. Molecular
phylogenetic approach for studying life-history evolution: The ambiguous
example of the genus medicago l. \emph{Proceedings of the Royal Society
of London B: Biological Sciences} \textbf{265}:1141--1151.

\leavevmode\hypertarget{ref-brown1996oldlist}{}%
Brown PM. 1996. OLDLIST: A database of maximum tree ages. \emph{Tree
rings, environment, and humanity Radiocarbon} \textbf{1996}:727--731.

\leavevmode\hypertarget{ref-charnov1973life}{}%
Charnov EL, Schaffer WM. 1973. Life-history consequences of natural
selection: Cole's result revisited. \emph{The American Naturalist}
\textbf{107}:791--793.

\leavevmode\hypertarget{ref-cole1954population}{}%
Cole LC. 1954. The population consequences of life history phenomena.
\emph{The Quarterly Review of Biology} \textbf{29}:103--137.

\leavevmode\hypertarget{ref-conti1999phylogenetic}{}%
Conti E, Soltis DE, Hardig TM, Schneider J. 1999. Phylogenetic
relationships of the silver saxifrages (saxifraga, sect. Ligulatae
haworth): Implications for the evolution of substrate specificity, life
histories, and biogeography. \emph{Molecular Phylogenetics and
Evolution} \textbf{13}:536--555.

\leavevmode\hypertarget{ref-corbin2004competition}{}%
Corbin JD, D'Antonio CM. 2004. Competition between native perennial and
exotic annual grasses: Implications for an historical invasion.
\emph{Ecology} \textbf{85}:1273--1283.

\leavevmode\hypertarget{ref-cruz2009molecular}{}%
Cruz-Mazo G, Buide M, Samuel R, Narbona E. 2009. Molecular phylogeny of
scorzoneroides (asteraceae): Evolution of heterocarpy and annual habit
in unpredictable environments. \emph{Molecular phylogenetics and
evolution} \textbf{53}:835--847.

\leavevmode\hypertarget{ref-dai2011drought}{}%
Dai A. 2011. Drought under global warming: A review. \emph{Wiley
Interdisciplinary Reviews: Climate Change} \textbf{2}:45--65.

\leavevmode\hypertarget{ref-daru2018widespread}{}%
Daru BH, Park DS, Primack RB, Willis CG, Barrington DS, Whitfeld TJ,
Seidler TG, Sweeney PW, Foster DR, Ellison AM, others. 2018. Widespread
sampling biases in herbaria revealed from large-scale digitization.
\emph{New Phytologist} \textbf{217}:939--955.

\leavevmode\hypertarget{ref-datson2008climate}{}%
Datson P, Murray B, Steiner K. 2008. Climate and the evolution of
annual/perennial life-histories in nemesia (scrophulariaceae).
\emph{Plant Systematics and Evolution} \textbf{270}:39--57.

\leavevmode\hypertarget{ref-davis2015herbarium}{}%
Davis CC, Willis CG, Connolly B, Kelly C, Ellison AM. 2015. Herbarium
records are reliable sources of phenological change driven by climate
and provide novel insights into species' phenological cueing mechanisms.
\emph{American Journal of Botany} \textbf{102}:1599--1609.

\leavevmode\hypertarget{ref-dittberner2018natural}{}%
Dittberner H, Korte A, Mettler-Altmann T, Weber A, Monroe G, Meaux J de.
2018. Natural variation in stomata size contributes to the local
adaptation of water-use efficiency in arabidopsis thaliana.
\emph{bioRxiv} 253021.

\leavevmode\hypertarget{ref-R-geiger_b}{}%
Eastman J, Alfaro M, Joyce P, Hipp A, Harmon L. 2011. A novel
comparative method for identifying shifts in the rate of character
evolution on trees. \emph{Evolution} \textbf{65}:3578--3589.

\leavevmode\hypertarget{ref-evans2005climate}{}%
Evans ME, Hearn DJ, Hahn WJ, Spangle JM, Venable DL. 2005. Climate and
life-history evolution in evening primroses (oenothera, onagraceae): A
phylogenetic comparative analysis. \emph{Evolution}
\textbf{59}:1914--1927.

\leavevmode\hypertarget{ref-franco1996life}{}%
Franco M, Silvertown J. 1996. Life history variation in plants: An
exploration of the fast-slow continuum hypothesis. \emph{Phil Trans R
Soc Lond B} \textbf{351}:1341--1348.

\leavevmode\hypertarget{ref-friedman2015all}{}%
Friedman J, Rubin MJ. 2015. All in good time: Understanding annual and
perennial strategies in plants. \emph{American journal of botany}
\textbf{102}:497--499.

\leavevmode\hypertarget{ref-gbifdownload}{}%
GBIF. 2018. GBIF occurrence download.

\leavevmode\hypertarget{ref-R-mvtnorm}{}%
Genz A, Bretz F. 2009. Computation of multivariate normal and t
probabilities, Lecture notes in statistics. Heidelberg: Springer-Verlag.

\leavevmode\hypertarget{ref-R-MCMCglmm}{}%
Hadfield JD. 2010. MCMC methods for multi-response generalized linear
mixed models: The MCMCglmm R package. \emph{Journal of Statistical
Software} \textbf{33}:1--22.

\leavevmode\hypertarget{ref-R-geiger_d}{}%
Harmon L, Weir J, Brock C, Glor R, Challenger W. 2008. GEIGER:
Investigating evolutionary radiations. \emph{Bioinformatics}
\textbf{24}:129--131.

\leavevmode\hypertarget{ref-R-logistf}{}%
Heinze G, Ploner M. 2018. Logistf: Firth's bias-reduced logistic
regression.

\leavevmode\hypertarget{ref-R-purrr}{}%
Henry L, Wickham H. 2018. Purrr: Functional programming tools.

\leavevmode\hypertarget{ref-R-raster}{}%
Hijmans RJ. 2018. Raster: Geographic data analysis and modeling.

\leavevmode\hypertarget{ref-R-phylolm}{}%
Ho LST, Ane C. 2014. A linear-time algorithm for gaussian and
non-gaussian trait evolution models. \emph{Systematic Biology}
\textbf{63}:397--408.

\leavevmode\hypertarget{ref-kerr2003space}{}%
Kerr JT, Ostrovsky M. 2003. From space to species: Ecological
applications for remote sensing. \emph{Trends in ecology \& evolution}
\textbf{18}:299--305.

\leavevmode\hypertarget{ref-kogan2016modelling}{}%
Kogan F, Guo W, Strashnaia A, Kleshenko A, Chub O, Virchenko O. 2016.
Modelling and prediction of crop losses from noaa polar-orbiting
operational satellites. \emph{Geomatics, Natural Hazards and Risk}
\textbf{7}:886--900.

\leavevmode\hypertarget{ref-kogan2001operational}{}%
Kogan FN. 2001. Operational space technology for global vegetation
assessment. \emph{Bulletin of the American Meteorological Society}
\textbf{82}:1949--1964.

\leavevmode\hypertarget{ref-lang2018using}{}%
Lang PL, Willems FM, Scheepens J, Burbano HA, Bossdorf O. 2018. Using
herbaria to study global environmental change. PeerJ Preprints.

\leavevmode\hypertarget{ref-lelievre2009current}{}%
Lelièvre F, Volaire F. 2009. Current and potential development of
perennial grasses in rainfed mediterranean farming systems. \emph{Crop
Science} \textbf{49}:2371--2378.

\leavevmode\hypertarget{ref-mandakova2012whole}{}%
Mandáková T, Mummenhoff K, Al-Shehbaz IA, Mucina L, Mühlhausen A, Lysak
MA. 2012. Whole-genome triplication and species radiation in the
southern african tribe heliophileae (brassicaceae). \emph{Taxon}
\textbf{61}:989--1000.

\leavevmode\hypertarget{ref-Mojica2016}{}%
Mojica JP, Mullen J, Lovell JT, Monroe JG, Paul JR, Oakley CG, McKay JK.
2016. Genetics of water use physiology in locally adapted Arabidopsis
thaliana. \emph{Plant Science}.
doi:\href{https://doi.org/10.1016/j.plantsci.2016.03.015}{10.1016/j.plantsci.2016.03.015}
